Automatic discovery of cell types and microcircuitry from neural connectomics

### 2015-01-03 docs/paper.tex
update docs post konrad writing

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists, including connectivity, cell body
  location and the spatial distribution of synapses, in a
  principled and probabilistically-coherent manner. We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  C. elegans.
  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

\section*{Introduction}
Emerging connectomics techniques \autocite{Morgan2013,Zador2012}
promise to quantify the location and connectivity of each neuron
within a tissue volume. These massive datasets will far exceed the
capacity of neuroanatomists to manually trace small circuits, thus
necessitating computational, quantitative, and automatic methods for
understanding neural circuit structure.  The impact of this kind of
high-throughput transition has been seen before -- rise of sequencing
techniques necessitated the development of novel computational methods
to understand genomic structure, ushering in  bioinformatics
as an independent discipline \autocite{Koboldt2013}.

@@

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more complex than traditional motifs,
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development processes.

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{Deriving circuitry and types from connectomics data a. As input we take the connectivity betwen cells (b), the distance between them (c), the depth of the cell bodies (d), and the depth profile of the syanpses (e). f. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synpatic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. f.) shows the cell connectivity matrix with cells of the
    same type grouped together. g.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. h. shows the probability of connection between two cell types
that very rarely connect -- there's a background ``base'' connection rate
to account for errors in data, but the probability is very low. 
We can plot this these types (i) to show the relationship and spatial
connections, and how probable various types are to connect to one
another. j.) The connectivity between our discovered types changes as a
function of distance. }

@@

We start with a model for connectivity, the infinite stochastic block
model (iSBM)\autocite{Kemp2006a,Xu2006}, which has been shown to
meaningfully cluster connection graphs while learning the number of
hidden groups, or types. We extend this approach by adding distance
dependence to model salient aspects of microcircuitry via logistic and
exponential distance-link functions.  We additionally model cell body
depth unimodally and synapse density profile multimodially (see
Methods for mathematical details).

@@

\subsection{Metrics}
We focus on two questions we can ask from the learned model: which
cells are of the same type? and given any two cells, what is the
probability of their connection?

To assess the accuracy of a clustering compared to that determined by
neuroanatomists, we employ the adjustded Rand index (ARI). ARI
measures the similarity of two clusterings. While there are no perfect
clustering comparison metrics (see supplemental for a discussion of
ARI's shortcomings) all suffer from the need to balance
``overclustering'' (calling a single type two different types) with
``merging'' (calling two distinct types a single type).

To test whether our model accurately captures the true structure of
the data, we learn the model on the data with a subset of the links
missing and then compute our predictive accuracy. We perform 10-way
cross validation on a given dataset, learn the resulting model, and
use that model to predict the missing synapses. Each potential
link between cells is assigned a probability, and we compute
the area under the curve (AUC) for the resulting ROC curve. 
\fxwarning{Add additional ARI} 
\section*{Results}

\subsection{Ground Truth Validation}
To validate our model, we performed a series of simulations to test if
the model can accurately recover the true underlying network structure
and cell type identity.  We thus simulate data for which we know the
correct structure and comparing the estimated structure based on the
algorithm (see methods) with the one we used for simulation. We find
that the model does a good job of recovering the correct number of
cell types, (fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  For comparison, existing infinite
stochastic block model assumes cell type alone matters, and thus finds
small neighborhoods of connected nodes (instead of global connectivity
patterns). The model converges relatively quickly to an estimate of
the most probable values for the cell types, which is enabled by using
a combination of simulated annealing and parallelized Markov chain
Monte Carlo (see methods for details). Thus we can apply our model to
simulated datasets with structure and scale similar to that of our
biological datasets and recover the known correct structure.

\subsection{Model Mismatch}
Our type-and-distance assumptions for our connectivty model are
simplifying, and reflect a general consensus about the organization of
some (but not all) types of neural tissue. To show where our model
performs poorly, we attempt connectivity inference on five sets of
synthetic data. Our goal is to show that our model does not find
``structure'' when there is none. \fxwarning{Fix this phrasing -- we want to say 'if your model thinks there are types, there probably are!}.

\begin{itemize}
\item Data with type-specific connectivity and distance dependence
\item Data with type-specific connectivity and no distance dependence
\item Data generated according to a latent-position clustering model,
  where each cell belongs to a type which is part of a cluster in a
  low-dimensional space \fxwarning{Cite this paper}
\item cell type is combinatorial -- there are three possible
  properties a cell can have, and each subset of properties influences
  connectivity to neighboring cells. 
\item cell type is a continuum -- each cell's type continuously varies
  from zero to one, and the probability of a connection between cells
  depends on this type.
\end{itemize}

We generate ten synthetic datasets based on each of these assumptions,
and perform 10-way cross validation to predict link accuracy as well
as ARI to the known (``true'') type, type spatial extent, and type
number.

\fxwarning{Add results}

@@

Initial connectomic efforts reconstruct only a small section of tissue
-- as a result, many connections to cells outside that tissue volume
will be lost. We performed simulations to check if 1. edge effects
could destroy spatial structure and if edge effects could introduce
artifical, spurious spatial structure.

\fxwarning{Add ``intrudoce spatial structure where there is none}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{comparison.ai}}
  \caption{\textbf{Visualizing type inference uncertainty.} Our fully Bayesian model gives a confidence estimate (posterior probability) that any two given cells are of the same type. In b.) we visualize that cell-cell coassignment matrix, showing the probability that cell i is of the same type as cell j on a range from 0.0 to 1.0. The block structure shows subsets of cells which are believed to all belong to the same type. For comparison, a.) shows the anatomist-defined type for each cell, grouped broadly into the coarse types identified in the previous figure. \textbf{Link vs cluster accuracy} c. A comparison of the
    predictive accuracy (area under the curve) for hand-labeled
    anatomical data, versus inclusion of additional sources of
    information, as well as the clustering accuracy. Note that our model sacrifies very little predictive accuracy for additional clustering accuracy. By comparison, conventional methods fail at one or both.  d. ) The spatial extent (in depth and area) of the
    types identified by humans and our various algorithmic
    approaches.}

>>>>

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists, including connectivity, cell body
  location and the spatial distribution of synapses, in a
  principled and probabilistically-coherent manner. We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  C. elegans and a historic microprocessor.
  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

\section*{Introduction}
Emerging connectomics techniques \autocite{Morgan2013,Zador2012}
promise to quantify the location and connectivity of each neuron
within a tissue volume. These massive datasets will far exceed the
capacity of neuroanatomists to manually trace small circuits, thus
necessitating computational, quantitative, and automatic methods for
understanding neural circuit structure.  The impact of this kind of
high-throughput transition has been seen before -- rise of sequencing
techniques necessitated the development of novel computational methods
to understand genomic structure, ushering in  bioinformatics
as an independent discipline \autocite{Koboldt2013}.
\fxwarning{cite background article, cite the paper from Yuste compare supervised unsupervised on morphology, cite seungs new article}

@@

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more complex than traditional motifs \fxwarning{Add citation of Science “motifs” paper},
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development processes.

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{Deriving circuitry and types from connectomics data a. As input we take the connectivity between cells (b), the distance between them (c), the depth of the cell bodies (d), and the depth profile of the synapses (e). f. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synaptic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. f.) shows the cell connectivity matrix with cells of the
    same type grouped together. g.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. h. shows the probability of connection between two cell types
that very rarely connect -- there's a background ``base'' connection rate
to account for errors in data, but the probability is very low. 
We can plot this these types (i) to show the relationship and spatial
connections, and how probable various types are to connect to one
another. j.) The connectivity between our discovered types changes as a
function of distance. }

@@

We start with a model for connectivity, the infinite stochastic block
model (iSBM)\autocite{Kemp2006a,Xu2006}, which has been shown to
meaningfully cluster connection graphs while learning the number of
hidden groups, or types. We extend this approach by adding distance
dependence to model salient aspects of microcircuitry via logistic and
exponential distance-link functions.  We additionally model cell body
depth unimodally and synapse density profile multimodally (see
Methods for mathematical details).

@@

\subsection{Metrics}
To evaluate the quality of the model fit, we need to use information that quantifies aspects of the data for which we have ground truth information. We focus on two aspects of performance. First, if the model works well, then the probability that a pair of neurons is of the same type should be high if the neurons actually are of the same type. Second, the model should assign a high probability of connection between two cells if they have a connection in the underlying data. We term these two factors clustering accuracy and link prediction accuracy.

To assess the accuracy of a clustering compared to that determined by
neuroanatomists, we employ the adjusted Rand index (ARI). ARI is an established metric which measures the similarity of two clusterings -- two identical clusterings will have an ARI of 1.0. There are no perfect clustering comparison metrics as all suffer from the need to balance
``overclustering'' (calling a single type two different types) with
``merging'' (calling two distinct types a single type).  See supplemental for a discussion of
ARI's shortcomings.  Despite its shortcomings we use the ARI as a simple measure of clustering accuracy as it quantitatively captures the difference between the types discovered and the types believed by neuroanatomists.

To assess the accuracy of the model for connections, we use link prediction
accuracy. If our model accurately captures the true structure of
the data, it should be good at predicting if a link exists.
We thus train the model on the data with a subset of the links
marked as unobserved and thus compute our predictive accuracy. We perform 10-way
cross validation on a given dataset, learn the resulting model, and
use that model to predict the missing synapses. Each potential
link between cells is assigned a probability, and we compute
the area under the curve (AUC) for the resulting ROC curve. An AUC of 1.0 means that we perfectly predict the presence and absence of the missing synapses. We use link prediction accuracy to quantify how good the model is at discovering the underlying connectivity.
\fxwarning{Add additional ARI}

\section*{Results}
We will first establish that our algorithm works properly and try to understand its properties using simulated data. Subsequently we will analyze in detail a dataset on the retina. Lastly we will briefly discuss the analysis of data from the worm \textit{c.elegans} and from an ancient microprocessor.

\subsection{Ground Truth Validation}
To validate our model, we performed a series of simulations to test if
the model can accurately recover the true underlying network structure
and cell type identity.  We thus simulate data for which we know the
correct structure and compare the estimated structure based on the
algorithm (see methods) with the one we used for simulation. We find
that the model does a good job of recovering the correct number of
cell types, (fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  For comparison, the infinite
stochastic block model (fig~\ref{fig:synthetic}a, b, c, black line) assumes that only cell type matters, and thus finds
small neighborhoods of connected nodes (instead of global connectivity
patterns). Our model converges relatively quickly (see supplemental) to an estimate of
the most probable values for the cell types, which is enabled by using
a combination of simulated annealing and parallelized Markov chain
Monte Carlo (see methods for details). Thus our model at least is promising to be applied to biological datasets.

\subsection{Model Mismatch}
To understand how our model would perform on real datasets that do not perfectly adhere to its generative model we next want to analyze its performance in such cases. To show where our model
performs poorly, we attempt connectivity inference on five sets of
synthetic data. Our type-and-distance assumptions for our connectivity model are
simplifying, and reflect a general consensus about the organization of
some (but not all) types of neural tissue. We assume that type is discrete (as opposed to the lpcm). We assume that each neuron only has one type (as opposed to the mixed membership model). We assume that connectivity is a function of distance and type (as opposed to the classic stochastic block model). Our goal is to show that our model does not find
``structure'' when there is none.

We generate ten synthetic datasets based on assumptions of other models and analyze our model’s performance on these datasets. More specifically we perform 10-way cross validation to predict link accuracy as well as ARI to the known (``true'') type, type spatial extent, and type number. We find that x. We find that y. We find that z. Our results thus strongly suggest even when the data is not generated according to our own generative model that our results are meaningful.

\fxwarning{Add results} 
\fxwarning{add spatial extent of types}

\fxwarn{add to figure comments}
\begin{itemize}
\item Data with type-specific connectivity and distance dependence
\item Data with type-specific connectivity and no distance dependence
\item Data generated according to a latent-position clustering model,
  where each cell belongs to a type which is part of a cluster in a
  low-dimensional space \fxwarning{Cite this paper}
\item cell type is combinatorial -- there are three possible
  properties a cell can have, and each subset of properties influences
  connectivity to neighboring cells. 
\item cell type is a continuum -- each cell's type continuously varies
  from zero to one, and the probability of a connection between cells
  depends on this type.
\end{itemize}

@@

Connectomic efforts so far have reconstructed only small sections of tissue at a time. Consequently, many connections to cells outside that tissue volume
will be lost. We are concerned that this selective elimination of connectivity along the boundary might give the appearance of distance-dependent connectivity when there is none. We thus performed simulations to check if edge effects
could destroy spatial structure and if edge effects could introduce
artificial, spurious spatial structure. We measure the degree to which the model infers distance dependent connectivity for two synthetic datasets -- one with distance-dependence and one without. We find that, if there is no distance-dependence, edge effects do not artificially introduce distance dependence. However, if the section we are examining is too small, our model can miss the distance dependence. Thus with respect to distance-dependent connectivity inference, our model errs on the side of caution. But we also find that for spatial extend that is similar to the currently available datasets the effects of this are quite limited.

\fxwarning{Add in a line on this figure to show where the existing data is by looking at old-school tracer reconstruction in the retina. CITE }

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{comparison.ai}}
  \caption{\textbf{Visualizing type inference uncertainty.} Our fully Bayesian model gives a confidence estimate (posterior probability) that any two given cells are of the same type. In b.) we visualize that cell-cell coassignment matrix, showing the probability that cell i is of the same type as cell j on a range from 0.0 to 1.0. The block structure shows subsets of cells which are believed to all belong to the same type. For comparison, a.) shows the anatomist-defined type for each cell, grouped broadly into the coarse types identified in the previous figure. \textbf{Link vs cluster accuracy} c. A comparison of the
    predictive accuracy (area under the curve) for hand-labeled
    anatomical data, versus inclusion of additional sources of
    information, as well as the clustering accuracy. Note that our model sacrifices very little predictive accuracy for additional clustering accuracy. By comparison, conventional methods fail at one or both.  d. ) The spatial extent (in depth and area) of the
    types identified by humans and our various algorithmic
    approaches.}
